---
title: "R part 1"
author: "Sihui Lin"
date: "2026-03-15"
output: pdf_document
---


# 1 install data
```{r setup, message=FALSE, warning=FALSE}
Sys.setenv(
  RICU_DATA_PATH = "/Users/shruaibylsh/Desktop/stats c160/stats-c160/mimic-iv-2.2"
)

library(ricu)
library(data.table)
library(lubridate)
library(comorbidity)

attach_src("miiv", assign_env = .GlobalEnv)
```


# Load raw tables via ricu
```{r load-tables, message=FALSE}
icustays      <- as.data.table(load_src("icustays", "miiv"))
admissions    <- as.data.table(load_src("admissions", "miiv"))
patients      <- as.data.table(load_src("patients", "miiv"))
omr           <- as.data.table(load_src("omr", "miiv"))
diagnoses_icd <- as.data.table(load_src("diagnoses_icd", "miiv"))
services      <- as.data.table(load_src("services", "miiv"))
```


# Step 1: Build first-ICU-stay cohort
```{r cohort}
setorder(icustays, subject_id, intime, stay_id)
cohort <- icustays[, .SD[1L], by = subject_id,
                   .SDcols = c("hadm_id", "stay_id", "intime", "outtime")]

admissions_keep <- admissions[, .(hadm_id, admittime, admission_type,
                                  admission_location, hospital_expire_flag)]
cohort <- admissions_keep[cohort, on = "hadm_id"]

patients_keep <- patients[, .(subject_id, anchor_age, anchor_year, gender)]
cohort <- patients_keep[cohort, on = "subject_id"]

cat("Cohort (first ICU stay per patient):", nrow(cohort), "stays\n")
```


# Step 2: BMI: binary (18.5-25, >25)
```{r bmi}
omr_bmi <- omr[result_name %in% c("BMI", "BMI (kg/m2)"),
               .(subject_id,
                 chartdate = as.IDate(chartdate),
                 bmi_raw   = suppressWarnings(as.numeric(result_value)))]
omr_bmi <- omr_bmi[!is.na(bmi_raw) & bmi_raw > 10 & bmi_raw < 100]

# Average multiple readings on the same calendar day
omr_bmi <- omr_bmi[, .(bmi_raw = mean(bmi_raw, na.rm = TRUE)),
                   by = .(subject_id, chartdate)]
setkey(omr_bmi, subject_id, chartdate)

# Rolling join: most recent BMI on or before ICU admission date.
# Rename intime to 'chartdate' so keys align; omr_bmi[lookup, roll=TRUE] picks
# the largest chartdate in omr_bmi that is <= the lookup date, per subject_id.
lookup <- cohort[, .(subject_id, hadm_id, stay_id, intime,
                     chartdate = as.IDate(intime))]
setkey(lookup, subject_id, chartdate)
bmi_matched <- omr_bmi[lookup, roll = TRUE]

bmi_matched[, bmi_binary := fcase(
  bmi_raw >= 18.5 & bmi_raw <= 25, 0L,   # normal weight
  bmi_raw  > 25,                   1L,   # overweight / obese
  default  = NA_integer_                 # underweight or missing
)]

n_before <- nrow(cohort)
bmi_matched <- bmi_matched[!is.na(bmi_binary)]
cat("Excluded (no pre-admission BMI or BMI < 18.5):",
    n_before - nrow(bmi_matched), "\n")
cat("Analytic sample after BMI filter:", nrow(bmi_matched), "\n")
cat("BMI group counts  (0 = normal weight  [18.5, 25],",
    "1 = overweight/obese  > 25):\n")
print(table(bmi_matched$bmi_binary))
```


# Step 4: SOFA at 24 h via ricu
```{r sofa}
sofa_raw <- load_concepts("sofa", "miiv")
sofa_dt  <- as.data.table(sofa_raw)
t_col    <- index_var(sofa_raw)

sofa_24h <- sofa_dt[get(t_col) == as.difftime(24L, units = "hours"),
                    .(sofa = sofa),
                    by = stay_id]
```


# Step 5: Age and sex
# Age = anchor_age + (year(admittime) - anchor_year).
# Sex = factor(gender). Both computed in the assemble chunk below.


# Step 6: Elixhauser comorbidity score
```{r elixhauser}
run_elix <- function(dx, map) {
  comorbidity(as.data.frame(dx), id = "hadm_id", code = "icd_code",
              map = map, assign0 = FALSE)
}

dx_cohort <- diagnoses_icd[hadm_id %in% cohort$hadm_id]
elix10 <- run_elix(dx_cohort[icd_version == 10, .(hadm_id = as.character(hadm_id), icd_code)], "elixhauser_icd10_quan")
elix9  <- run_elix(dx_cohort[icd_version == 9,  .(hadm_id = as.character(hadm_id), icd_code)], "elixhauser_icd9_quan")

flag_cols <- setdiff(names(elix10), "hadm_id")
elix_all  <- rbindlist(list(as.data.table(elix10), as.data.table(elix9)), fill = TRUE)

# Column-wise max per hadm_id: 1 wins over 0 across ICD-9 and ICD-10
elix_agg <- elix_all[, lapply(.SD, function(x) max(x, na.rm = TRUE)),
                     by = hadm_id, .SDcols = flag_cols]

for (col in flag_cols) {
  idx <- which(is.infinite(elix_agg[[col]]))
  if (length(idx)) set(elix_agg, idx, col, 0L)
}

# score() requires a 'comorbidity' S3 object — restore class and map attr
# stripped by rbindlist/lapply before scoring.
elix_agg_df <- as.data.frame(elix_agg)
class(elix_agg_df)       <- c("comorbidity", "data.frame")
attr(elix_agg_df, "map") <- "elixhauser_icd10_quan"

elix_scores_vec <- score(elix_agg_df, weights = "vw", assign0 = FALSE)
elix_scores <- data.table(hadm_id    = as.integer(elix_agg$hadm_id),
                          elix_score = elix_scores_vec)
cat("Elixhauser score summary:\n")
print(summary(elix_scores$elix_score))
```


# Step 7: Admission type (medical / elective_surgical / nonelective_surgical)
```{r adm-type}
adm_info <- cohort[, .(hadm_id, admission_type, admission_location)]

adm_info[, elective := fifelse(
  admission_type %in% c("AMBULATORY OBSERVATION", "ELECTIVE") |
    (admission_type == "DIRECT OBSERVATION" &
       admission_location %in% c("PROCEDURE SITE", "PACU",
                                 "AMBULATORY SURGERY TRANSFER")) |
    (admission_type == "OBSERVATION ADMIT" &
       admission_location %in% c("PROCEDURE_SITE", "PACU")) |
    (admission_type == "SURGICAL SAME DAY ADMISSION" &
       admission_location %in% c("PROCEDURE_SITE", "PACU",
                                 "PHYSICIAN REFERRAL")),
  1L, 0L
)]

surgical_svcs <- c("CSURG", "NSURG", "ORTHO", "SURG", "TSURG",
                   "VSURG", "GYN", "OBS")

svc_cohort <- services[hadm_id %in% cohort$hadm_id,
                       .(hadm_id, transfertime, curr_service)]
setorder(svc_cohort, hadm_id, transfertime)
first_svc <- svc_cohort[, .SD[1L], by = hadm_id, .SDcols = "curr_service"]

adm_type_dt <- adm_info[first_svc, on = "hadm_id"]
adm_type_dt[, adm_type := fcase(
  !(curr_service %in% surgical_svcs),               "medical",
  curr_service %in% surgical_svcs & elective == 1L, "elective_surgical",
  curr_service %in% surgical_svcs & elective == 0L, "nonelective_surgical",
  default = NA_character_
)]
adm_type_dt[, adm_type := factor(adm_type,
                                  levels = c("medical", "elective_surgical",
                                             "nonelective_surgical"))]
adm_type_dt <- adm_type_dt[, .(hadm_id, adm_type)]

cat("Admission type distribution:\n")
print(table(adm_type_dt$adm_type, useNA = "ifany"))
```


# Assemble final dataset dat
```{r assemble}
dat <- copy(bmi_matched[, .(subject_id, hadm_id, stay_id, intime,
                             bmi_binary, bmi_raw)])

mort_map <- cohort[, .(hadm_id, hospital_expire_flag)]
dat[mort_map, on = "hadm_id", death := as.integer(i.hospital_expire_flag)]

dat[sofa_24h, on = "stay_id", sofa := i.sofa]

demog_map <- cohort[, .(hadm_id, admittime, anchor_age, anchor_year, gender)]
dat[demog_map, on = "hadm_id",
    `:=`(admittime   = i.admittime,
         anchor_age  = i.anchor_age,
         anchor_year = i.anchor_year,
         gender      = i.gender)]

dat[, age := anchor_age + (year(admittime) - anchor_year)]
dat[, sex := factor(gender, levels = c("M", "F"))]
dat[, c("admittime", "anchor_age", "anchor_year", "gender") := NULL]

dat[elix_scores,  on = "hadm_id", elix_score := i.elix_score]
dat[adm_type_dt,  on = "hadm_id", adm_type   := i.adm_type]
dat[, adm_type := factor(adm_type, levels = c("medical", "elective_surgical",
                                               "nonelective_surgical"))]

setcolorder(dat, c("subject_id", "hadm_id", "stay_id", "intime",
                   "bmi_binary", "bmi_raw",
                   "death", "sofa",
                   "age", "sex", "elix_score", "adm_type"))

cat("Final dat:", nrow(dat), "rows x", ncol(dat), "cols\n")
cat("Columns:", paste(names(dat), collapse = ", "), "\n")
```


# Summary statistics for dat
```{r summary}
cat("=== nrow(dat) ===\n")
print(nrow(dat))

cat("\n=== sum(is.na(.)) per column ===\n")
print(sapply(dat, function(x) sum(is.na(x))))

cat("\n=== table(dat$bmi_binary)  [0 = normal, 1 = overweight/obese] ===\n")
print(table(dat$bmi_binary, dnn = "bmi_binary"))

cat("\n=== table(dat$adm_type) ===\n")
print(table(dat$adm_type, useNA = "ifany"))

cat("\n=== mean(dat$death) ===\n")
print(mean(dat$death, na.rm = TRUE))
```
